Reconstructing the Lensing Mass in the Universe \\ from Photometric Catalog Data

### 2012-10-30 doc/paper-lcr.tex
fixed refs, citations

\section{Introduction}

@@

Several attempts have been made to do better than simply constructing a
large sample of objects, and averaging over the resulting convergence
distribution. The weak lensing effect can be detected observationally 
by measuring the
small distortions it induces on the images of many background galaxies
in the field \citep[see e.g.][for a review]{SaasFee2006}. In a 30''
radius aperture around the strong lens Q0957$+$561,
\citet{NakajimaEtal2009} used deep HST imaging to measure a weak lensing
convergence of $0.17/pm0.06$, which was used in the time delay distance
measurement of \citep{FadelyEtal2009}. While this system has an
unusually high external convergence due to its location on the edge of a
cluster of galaxies, the precision with which the convergence at the
lens position can be estimated is indicative of what can be achieved
with high quality weak lensing shear data. We note that this uncertainty
is not dependent on the value of the convergence, but instead depends
primarily on the number density of weakly lensed, measurable galaxies
available.

@@

The \MS results have been used to analyse real observations as well. 
\citet{SuyuEtal2010} selected \MS lines of sight by their apparent
galaxy overdensity, in a 45 arcsec radius aperture down to $i < 24.5$,
to match the observed overdensity towards the time delay lens
B1608$+$656 \citep{FassnachtEtal2011}.  The resulting distribution of
$\kappa$ values from the ray-tracing was taken to be an estimate of 
$\Pr(\kappa)$, which was then marginalised over when inferring the time
delay distance in this system. 
% B1608$+$656 is indeed a high $\kappa$
% line of sight, but it does not reside in a massive cluster environment.
In a companion paper to this one, \citet{GreeneEtal2012} investigated
improvements to this method by weighting the galaxy counts by galaxy
mass, and perpendicular distance to the line of sight, again using the
\MS mock catalogs and their associated $\kappa$ values to construct
$\Pr(\kappa|\data)$.

In this paper, we combine the halo model reconstruction approach of
\citeauthor{GunnarssonEtal2006} and \citeauthor{WongEtal2011}, with the
idea of calibrating to simulations from \citeauthor{SuyuEtal2009}, in
order to put the halo model for  estimating the weak lensing convergence
due to mass along the line of sight to a distant source into a sound
framework for evaluating this important source of systematic
uncertainty, as we look forward to ever increasing sample sizes and
statistical precision. Taking the $\kappa$ values from
\citeauthor{HilbertEtal2009}'s ray-tracing calculations as the ``truth''
that we have to recover, we use the \MS mock galaxy catalogs first to
{\it calibrate} the reconstruction, and account for unseen mass and
voids, and then to {\it test} the accuracy of the line of sight mass
reconstruction under this assumption. Probabilistically assigning mass
and redshift to every observed galaxy in a given observed field, we
generate Monte Carlo sample line of sight mass distributions, and so
construct the PDF $\Pr(\kappa|\data)$ for that field. We then emulate
the combination of many such PDFs to quantify the residual biases that
would be translated to the global (including cosmological) parameters.
In doing so, we aim to answer the following questions:

@@

The relationship between convergence $\kappa$ and magnification $\mu$ is
\bea 
\mu &=&       \frac{1}{(1-\kappa)^2 - |\gamma|^2}, \notag \\
    &\approx& \frac{1}{(1-2\kappa)}
\label{eq:MassSheet:mag}
\eea
where we have used the fact that the magnitude of the gravitational
shear at any point is typically comparable to the magnitude of the 
convergence. For small $\kappa$, the fractional uncertainty in the
inferred luminosity of a magnified source will be approximately  twice
the absolute uncertainty in the convergence~$\kappa$. These
relationships will allow us to put our results on convergence estimation
in context.

@@

The projected surface mass density $\Sigma(r)$ for this mass profile is derived in \citet{BMO}. Each halo in our catalogue then contributes to the line of sight convergence by
\be
\label{eq:kappai}
\kappa_i =\Sigma_{i}/\Sigma_{\mathrm{cr}}(z_i),
\ee
 where $\Sigma_{\rm cr}(z)$ is the critical surface density given by:
\be \label{eq:sigcrit} 
\Sigma_{\rm cr}(z) \equiv \frac{c^2 D_{\rm os}}{4 \pi G D_{\rm od} D_{\rm ds}},
\ee
 the D's are the angular diameter distances between the objects refered to in the subscripts: o is the observer, d is the deflector and s is the source. We take the total convergence from all the halos along the line of sight to be given by
\be \label{eq:kappasummu}
\kappah=\sum_{i}\kappa_i.
\ee

To estimate a halo's contribution to $\kappah$ we must first know the halo's mass.
For the calibration sightlines of the \MS halo masses are known, but for 
real lines of sight this can only be infered from the stellar masses of the
galaxies within a halo. We use the relation of \citet{BehrooziEtal2009} to generate 
halo masses from stellar masses; the details of this proceedure are outlined in
Appendix \ref{appendix:MSMH}.

@@

Before investigating the uncertainties caused by imperfect knowledge of
halo mass and redshift, we investigate the uncertainties induced by a
magnitude limited reconstruction and a field of view limited
reconstruction. The halos in our catalogue are populated by galaxies and
given magnitudes according to the semi-analytic model of
\citet{deLucia+Blaizot2007} and by applying magnitude cuts to our
catalogue, we can investigate the scatter caused by unobserved halos.
The majority of the convergence comes from halos with an $i$ magnitude
between 18 and 24. Figure \ref{fig:where} shows the cumulative
contribution to $\kappah$ as a function of projected distance from the
line of sight, magnitude and redshift. Unsurprisingly we find that most
of the convergence comes from halos whose galaxies are close to the line
of sight. As a function of magnitude, we find that $\kappah$ is
dominated by objects with magnitudes between $i=18$ and $i=24$. Objects
brighter than $i=18$ are either too rare or too close to the observer to
make a significant contribution to the convergence. Objects fainter than
$i=24$ are too small to be important, unless they are extremely close to
the line of sight, but we find that including including halos fainter
than $i=24$ does not improve our reconstruction. Figure \ref{fig:magcut}
shows the scatter on $\kappah-\kappa$ (where $\kappah$ is shifted so
that it's mean is zero) as a function of reconstruction magnitude limit.
Since these objects must be extremely close to the line of sight to make
a non-zero $\kappah$ contribution it is likely that neglecting stellar
mass and using a spherical NFW prescription are too naive to adequately
reconstruct $\kappax$; thus we do not expect a deeper survey to decrease
the size of the uncertainty in mapping $\kappah$ onto $\kappa$. We find
that halos at all redshifts contribute to the convergence, but halos at
redshifts of $z \sim z_{\rm source}/2$ contribute the most.

@@

We generate a stellar mass for each halo according to the stellar mass--halo mass relation of \citet{BehrooziEtal2009}. From this stellar mass we simulate 
an observed stellar mass by drawing a sample from $\pr(\log(M_{* \mathrm {obs}})|\log(M_{* \mathrm {true}}))$ which is given by a Gaussian of width $\sigma_{M_*}$ centred on $\log(M*_{\mathrm {true}})$. Where a spectroscopic redshift exists stellar masses can be estimated with typical uncertainties of 0.15 dex \citep{AugerEtal2009}, however with photometric redshifts stellar mass uncertainties are typically three times as large; we use $\sigma_{M_*}=0.15$ for halos with a spectroscopic redshift and $\sigma_{M_*}=0.45$ otherwise. For photometric redshift errors we draw a redshift from $\pr(z_{\rm true}|z_{\rm obs})$ which we take as a Gaussian of width $0.1(1+z_{\rm spec})$ centred on $z_{\rm spec}$, where $z_{\rm spec}$ is the halo's true redshift in the \MS catalogue.

Given an uncertain stellar mass and redshift it is possible to infer a halo mass using the stellar mass--halo mass relation of \citet{BehrooziEtal2009}. This proceedure requires an inversion of the relation given in \citet{BehrooziEtal2009} and correctly inverting the relation's uncertainties requires care: the proceedure we use to do this is given in Appendix \ref{appendix:MSMH}. By drawing sample halo masses and redshifts, we can infer a sample $\kappah$ using the proceedure of Section \ref{sec:model}; repeatedly drawing samples allows us to estimate P$(\kappah)$ for each reconstructed line of sight.

@@

\section{Testing for Systematic errors}
\label{sec:biases}
\begin{figure*}
\includegraphics[width=\textwidth]{figs/biasplots.eps}
\caption[magcut]{Estimating $\kappa-\kappa_{\rm true}$ from subsets of
lenses. By subtracting the true (ray-traced) convergence from the
infered P($\kappa$) for each line of sight we are left with a PDF that
should be estimating zero, these can be multiplied together for multiple
lines of sight to test for a possible bias. These plots show the
expectation value of $\prod_{i=1}^N \pr_i(\kappa-\kappa_{\rm true})$ -
deviations from zero can be seen as a bias. The solid, dashed and dotted
lines are for combining 20, 5 and 2 lenses respectively. Green lines are
from multiplying P($\kappa$) infered from photometry alone, whilst red
lines are from multiplying P($\kappa$) infered from photometry and a
spectroscopic redshift for all halos with $i<23$ and all halos within 1
arcminute and $i<24$. The left most plot shows randomly selected lines
of sight. The central plot shows sight lines randomly selected from the
third of lines whose P($\kappa$) is most tightly constrained. The right
hand plot shows only lines of sight with external shear of 0.03 or
greater.}
\label{fig:where}
\end{figure*}

@@

\subsubsection{Selecting only high shear lines of sight}
\todo{phil}{Write up why lenses are likely to be gamma biased}
\notes{The uncertainties from modeling a strong lens system depends on the comfiguration of the source images. With only two point images ofthe source the lens slope and mass are degenerate; the mass-slope degeneracy\citep{mass-slope degeneracy} cite saha?. Where lensed arcs or four point images are present the mass-slope degeneracy can be broken and cosmology can be extracted, \citep{Suyu2012a}. The detectability of arcs depend on the source, but the presence of four images (so called quads) depend on the shear of the lens model - a high shear is needed to produce a quad. The shear is almost certainly dominated by the lens model itself, but it is possible that quads are biased towards high external shear.}

@@

TC thanks Vasily Belokurov for guidance and discussions.
We thank Risa Wechsler and Peter Behroozi for useful discussions and 
suggestions.\notes{do we?}
\input{acknowledgments.tex}

>>>>

\section{Introduction}
\label{sec:intro}

@@

Several attempts have been made to do better than simply constructing a
large sample of objects, and averaging over the resulting convergence
distribution. The weak lensing effect can be detected observationally 
by measuring the
small distortions it induces on the images of many background galaxies
in the field \citep[see e.g.][for a review]{Schneider2006}. In a 30''
radius aperture around the strong lens Q0957$+$561,
\citet{NakajimaEtal2009} used deep HST imaging to measure a weak lensing
convergence of $0.17\pm0.06$, which was used in the time delay distance
measurement of \citep{FadelyEtal2009}. While this system has an
unusually high external convergence due to its location on the edge of a
cluster of galaxies, the precision with which the convergence at the
lens position can be estimated is indicative of what can be achieved
with high quality weak lensing shear data. We note that this uncertainty
is not dependent on the value of the convergence, but instead depends
primarily on the number density of weakly lensed, measurable galaxies
available.

@@

The \MS results have been used to analyse real observations as well. 
\citet{SuyuEtal2010} selected \MS lines of sight by their apparent
galaxy overdensity, in a 45 arcsec radius aperture down to $i < 24.5$,
to match the observed overdensity towards the time delay lens
B1608$+$656 \citep{FassnachtEtal2011}.  The resulting distribution of
$\kappa$ values from the ray-tracing was taken to be an estimate of 
$\Pr(\kappa)$, which was then marginalised over when inferring the time
delay distance in this system. 
% B1608$+$656 is indeed a high $\kappa$
% line of sight, but it does not reside in a massive cluster environment.
In a companion paper to this one, Greene et al (submitted) investigate
improvements to this method by weighting the galaxy counts by galaxy
mass, and perpendicular distance to the line of sight, again using the
\MS mock catalogs and their associated $\kappa$ values to construct
$\Pr(\kappa|\data)$.

In this paper, we combine the halo model reconstruction approach of
\citeauthor{GunnarssonEtal2006} and \citeauthor{WongEtal2011}, with the
idea of calibrating to simulations from \citeauthor{SuyuEtal2010}, in
order to put the halo model for  estimating the weak lensing convergence
due to mass along the line of sight to a distant source into a sound
framework for evaluating this important source of systematic
uncertainty, as we look forward to ever increasing sample sizes and
statistical precision. Taking the $\kappa$ values from
\citeauthor{HilbertEtal2009}'s ray-tracing calculations as the ``truth''
that we have to recover, we use the \MS mock galaxy catalogs first to
{\it calibrate} the reconstruction, and account for unseen mass and
voids, and then to {\it test} the accuracy of the line of sight mass
reconstruction under this assumption. Probabilistically assigning mass
and redshift to every observed galaxy in a given observed field, we
generate Monte Carlo sample line of sight mass distributions, and so
construct the PDF $\Pr(\kappa|\data)$ for that field. We then emulate
the combination of many such PDFs to quantify the residual biases that
would be translated to the global (including cosmological) parameters.
In doing so, we aim to answer the following questions:

@@

The relationship between convergence $\kappa$ and magnification $\mu$ is
\bea 
\mu &=&       \frac{1}{(1-\kappa)^2 - |\gamma|^2}, \notag \\
    &\approx& (1+2\kappa)
\label{eq:MassSheet:mag}
\eea
where we have used the facts that the magnitude of the gravitational
shear at any point is typically comparable to the magnitude of the 
convergence, and that both are typically small. 
Under these assumptions, the fractional uncertainty in the
inferred luminosity of a magnified source will be approximately  twice
the absolute uncertainty in the convergence~$\kappa$. These
relationships will allow us to put our results on convergence estimation
in context.

@@

The projected surface mass density $\Sigma(r)$ for this mass profile is derived in \citet{BMO}. Each halo in our catalogue then contributes to the line of sight convergence by
\be
\label{eq:kappai}
\kappa_i =\Sigma_{i}/\Sigma_{\mathrm{cr}}(z_i,\zs),
\ee
where the critical surface density was defined in \Eref{eq:sigcrit}.
Following the first order approximation of \citet{HilbertEtal2009}
outlined in \Sref{sec:MS} above, we compute 
the total convergence from all the halos along the line
of sight using
\be 
\label{eq:kappasummu}
\kappah = \sum_{i} \kappa_i.
\ee

To estimate a halo's contribution to $\kappah$ we must first know the halo's mass.
For the calibration sightlines of the \MS halo masses are known, but for 
real lines of sight this can only be infered from the stellar masses of the
galaxies within a halo. We use the relation of \citet{BehrooziEtal2010} to generate 
halo masses from stellar masses; the details of this proceedure are outlined in
Appendix \ref{appendix:MSMH}.

@@

Before investigating the uncertainties caused by imperfect knowledge of
halo mass and redshift, we investigate the uncertainties induced by a
magnitude limited reconstruction and a field of view limited
reconstruction. The halos in our catalogue are populated by galaxies and
given magnitudes according to the semi-analytic model of
\citet{DeLucia+Blaizot2007} and by applying magnitude cuts to our
catalogue, we can investigate the scatter caused by unobserved halos.
The majority of the convergence comes from halos with an $i$ magnitude
between 18 and 24. \Fref{fig:where} shows the cumulative
contribution to $\kappah$ as a function of projected distance from the
line of sight, magnitude and redshift. Unsurprisingly we find that most
of the convergence comes from halos whose galaxies are close to the line
of sight. As a function of magnitude, we find that $\kappah$ is
dominated by objects with magnitudes between $i=18$ and $i=24$. Objects
brighter than $i=18$ are either too rare or too close to the observer to
make a significant contribution to the convergence. Objects fainter than
$i=24$ are too small to be important, unless they are extremely close to
the line of sight, but we find that including including halos fainter
than $i=24$ does not improve our reconstruction. Figure \ref{fig:magcut}
shows the scatter on $\kappah-\kappa$ (where $\kappah$ is shifted so
that it's mean is zero) as a function of reconstruction magnitude limit.
Since these objects must be extremely close to the line of sight to make
a non-zero $\kappah$ contribution it is likely that neglecting stellar
mass and using a spherical NFW prescription are too naive to adequately
reconstruct $\kappax$; thus we do not expect a deeper survey to decrease
the size of the uncertainty in mapping $\kappah$ onto $\kappa$. We find
that halos at all redshifts contribute to the convergence, but halos at
redshifts of $z \sim z_{\rm source}/2$ contribute the most.

@@

We generate a stellar mass for each halo according to the stellar mass--halo mass relation of 
\citet{BehrooziEtal2010}. From this stellar mass we simulate 
an observed stellar mass by drawing a sample from $\pr(\log(M_{* \mathrm {obs}})|\log(M_{* \mathrm {true}}))$ which is given by a Gaussian of width $\sigma_{M_*}$ centred on $\log(M*_{\mathrm {true}})$. Where a spectroscopic redshift exists stellar masses can be estimated with typical uncertainties of 0.15 dex \citep{AugerEtal2009}, however with photometric redshifts stellar mass uncertainties are typically three times as large; we use $\sigma_{M_*}=0.15$ for halos with a spectroscopic redshift and $\sigma_{M_*}=0.45$ otherwise. For photometric redshift errors we draw a redshift from $\pr(z_{\rm true}|z_{\rm obs})$ which we take as a Gaussian of width $0.1(1+z_{\rm spec})$ centred on $z_{\rm spec}$, where $z_{\rm spec}$ is the halo's true redshift in the \MS catalogue.

Given an uncertain stellar mass and redshift it is possible to infer a
halo mass using the stellar mass--halo mass relation of
\citet{BehrooziEtal2010}. This proceedure requires an inversion of the
relation given in \citet{BehrooziEtal2010} and correctly inverting the
relation's uncertainties requires care: the proceedure we use to do this
is given in Appendix \ref{appendix:MSMH}. By drawing sample halo masses
and redshifts, we can infer a sample $\kappah$ using the proceedure of
Section \ref{sec:model}; repeatedly drawing samples allows us to
estimate P$(\kappah)$ for each reconstructed line of sight.

@@

\section{Testing for Systematic errors}
\label{sec:biases}
\begin{figure*}
\includegraphics[width=\textwidth]{figs/biasplots.eps}
\caption[magcut]{Estimating $\kappa-\kappa_{\rm true}$ from subsets of
lenses. By subtracting the true (ray-traced) convergence from the
infered P($\kappa$) for each line of sight we are left with a PDF that
should be estimating zero, these can be multiplied together for multiple
lines of sight to test for a possible bias. These plots show the
expectation value of $\prod_{i=1}^N \pr_i(\kappa-\kappa_{\rm true})$ -
deviations from zero can be seen as a bias. The solid, dashed and dotted
lines are for combining 20, 5 and 2 lenses respectively. Green lines are
from multiplying P($\kappa$) infered from photometry alone, whilst red
lines are from multiplying P($\kappa$) infered from photometry and a
spectroscopic redshift for all halos with $i<23$ and all halos within 1
arcminute and $i<24$. The left most plot shows randomly selected lines
of sight. The central plot shows sight lines randomly selected from the
third of lines whose P($\kappa$) is most tightly constrained. The right
hand plot shows only lines of sight with external shear of 0.03 or
greater.}
\label{fig:biasplots}
\end{figure*}

@@

\subsubsection{Selecting only high shear lines of sight}
\todo{phil}{Write up why lenses are likely to be gamma biased}
\notes{The uncertainties from modeling a strong lens system depends on
the comfiguration of the source images. With only two point images ofthe
source the lens slope and mass are degenerate; the mass-slope
degeneracy\citep[need citation][Saha?]{mass-slope degeneracy}. Where lensed arcs or
four point images are present the mass-slope degeneracy can be broken
and cosmology can be extracted, \citep{Suyu2012a}. The detectability of
arcs depend on the source, but the presence of four images (so called
quads) depend on the shear of the lens model - a high shear is needed to
produce a quad. The shear is almost certainly dominated by the lens
model itself, but it is possible that quads are biased towards high
external shear.}

@@

We thank Vasily Belokurov, Risa Wechsler and Peter Behroozi 
for useful discussions and suggestions.
\input{acknowledgments.tex}

### 2012-10-30 doc/references.tex
more intro in hilbert 2011 etc, MS section added
fixed refs, citations

\bibitem[\protect\citeauthoryear{Behroozi, Conroy, 
\& Wechsler}{2010}]{BehrooziEtal2009} Behroozi P.~S., Conroy C., Wechsler R.~H., 2010, ApJ, 717, 379

@@

\bibitem[\protect\citeauthoryear{De Lucia 
\& Blaizot}{2007}]{deLucia+Blaizot2007} De Lucia G., Blaizot J., 2007, MNRAS, 375, 2

@@

@@

@@

>>>>

\bibitem[\protect\citeauthoryear{Behroozi, Conroy, 
\& Wechsler}{2010}]{BehrooziEtal2010} Behroozi P.~S., Conroy C., Wechsler R.~H., 2010, ApJ, 717, 379

@@

\bibitem[\protect\citeauthoryear{De Lucia 
\& Blaizot}{2007}]{DeLucia+Blaizot2007} De Lucia G., Blaizot J., 2007, MNRAS, 375, 2

@@

\bibitem[Gunnarsson et al.(2006)]{GunnarssonEtal2006} Gunnarsson, C., 
Dahl{\'e}n, T., Goobar, A., J{\"o}nsson, J., M{\"o}rtsell, E.\ 2006, \apj, 640, 417

@@

\bibitem[Hilbert et al.(2011)]{HilbertEtal2011} Hilbert, S., Gair, 
J.~R., \& King, L.~J.\ 2011, \mnras, 412, 1023

@@

\bibitem[Schneider(2006)]{Schneider2006} Schneider, P.\ 2006, 
in Saas-Fee Advanced Course 33: Gravitational Lensing: Strong, Weak and Micro, 
269
